Geospatial and Temporal Analysis of Toronto's Cycling Infrastructure

### 2024-09-25 paper/paper.qmd
paper second draft

---
title: "Geospatial and Temporal Analysis of Toronto's Cycling Infrastructure"
subtitle: "A Data-Driven Approach"
author: 
  - Ziheng Zhong
thanks: "Code and data are available at: https://github.com/iJustinn/Toronto_Cycling_Network"
date: today
date-format: long
abstract: "This paper explores the development of Toronto's cycling infrastructure. Through a comprehensive analysis of installation dates, upgrades, and comfort levels of bike lanes, this study highlights the trends in cycling infrastructure expansion from 2001 to 2024. A significant focus is placed on the comfort levels of different types of lanes, offering insights into how the infrastructure meets the needs of cyclists. The spatial distribution of bike lanes is also analyzed, revealing discrepancies in access between downtown and suburban areas. The findings emphasize the need for continued investment in both the quality and expansion of cycling infrastructure to promote sustainable transportation in Toronto."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# Load packages
package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr", "here", "knitr", "kableExtra")

# Function to check and install missing packages
install_and_load <- function(package_list) {
  for (package in package_list) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

install_and_load(package_list)
```

@@

Urban cycling plays a critical role in advancing sustainable transportation strategies, especially in rapidly growing cities like Toronto. With the city facing increasing congestion, promoting active transportation modes such as cycling is essential for reducing greenhouse gas emissions, improving public health, and enhancing urban mobility. Since the early 2000s, Toronto has made significant strides in expanding its cycling infrastructure, creating a network of bike lanes, multi-use paths, and shared roadways to support cyclists citywide.

This paper analyzes the evolution of Toronto's cycling infrastructure over the past two decades, focusing on the installation and upgrade trends of bikeways and their classification by comfort level. Using the 'Cycling Network' dataset from Open Data Toronto, the study examines both the spatial distribution of bike lanes and the varying levels of cycling comfort across different regions. The analysis highlights discrepancies between downtown and suburban areas, showcasing progress while identifying regions that remain underserved.

The study's findings underscore the need for continued investment in both the expansion and quality of Toronto's cycling infrastructure to encourage more widespread adoption of cycling as a viable mode of transportation. By addressing both the geographic imbalances and comfort levels of bike lanes, the city can better meet its sustainability goals.

The structure of this paper is as follows: the data sources and methodologies are presented in @sec-data, followed by an analysis of the results in @sec-res, and a discussion of the findings in @sec-dis.

@@

## Source
The dataset analyzed in this study, the Cycling Network dataset [@citeCyclingNetwork], was retrieved from Open Data Toronto [@citeODT]. This dataset is publicly available and is continually updated to reflect Toronto’s ongoing efforts to develop and maintain its cycling infrastructure. It contains geospatial and temporal information about various cycling-related features, including the types of bikeways (dedicated lanes, shared roads, multi-use trails), installation years, and any subsequent upgrades. By offering a detailed history of Toronto's cycling network development, this dataset plays a vital role in informing both urban planning policies and research into sustainable transportation options.

While the Cycling Network dataset was the primary source of data, other relevant datasets were explored, such as the Road Network File [@citeStatCan] from Statistics Canada, which provides detailed geospatial data on all roadways in Toronto. Another such alternative was the Major City-Wide Cycling Routes dataset [@citeCyclingRoutes], also available on Open Data Toronto [@citeODT]. However, this dataset focuses specifically on the large-scale cycling routes connecting different parts of the city, emphasizing connectivity at a macro level. While these two dataset are useful for analyzing high-level transportation patterns, it lacks the detailed, segment-level data—such as comfort levels, installation dates, and upgrade history—that are crucial for a deep analysis of cycling infrastructure. Since this paper focuses on both the growth and quality of cycling infrastructure, the Cycling Network dataset, with its more comprehensive coverage of individual bikeway segments, was a better fit for the goals of this study.

Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages which will be mentioned later in the @sec-data-method.

@@

The 'Cycling Network' dataset from Open Data Toronto tracks various aspects of Toronto's cycling infrastructure, such as installation dates, upgrades, street names, and the type of infrastructure (e.g., sharrows, multi-use trails). Each entry represents a segment of the cycling network, where real-world phenomena, such as the construction or upgrading of cycling paths, are documented. For example, when a cycling path is installed or upgraded, the responsible authorities collect information such as the installation year (e.g., 2001) and any subsequent upgrades (e.g., 2021). This data is then digitized into structured entries in the dataset.

To go from a physical event (e.g., the installation of a bike lane) to a dataset entry, detailed records are maintained by the city's transportation department. These records are geospatially coded (column geometry in this dataset), allowing each cycling path to be mapped precisely in relation to other city infrastructure. Thus, the phenomena of constructing or updating a bike lane becomes an entry with attributes such as the street name, road class, and installation history. This allows researchers to analyze trends in cycling infrastructure and its expansion over time.

@@

As mentioend above, the dataset contains several key variables that were used in this analysis:

@@

#### Comfort Level
Each bikeway is assigned a comfort level—high, moderate, or low—based according to the recorded Bikeway Type. These classification are self-defined, reling on factors such as physical separation from traffic and lane width of each Bikeway Type, detail can be found in @sec-app-comfort. This variable is central to understanding the usability of the infrastructure.

@@

Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR]. Also with support of additional packages in R: `sf` [@citeSf], `readr` [@citeReadr], `ggplot2` [@citeGgplot2], `osmdata` [@citeOsmdata], `tidyverse` [@citeTidyverse], `jsonlite` [@citeJsonlite], `dplyr` [@citeDplyr], `here` [@citeHere], `knitr` [@citeKnitr], `kableExtra` [@citeKableExtra].

@@

In the first data cleaning process, the focus was on extracting and cleaning geographical coordinates from the 'geometry' column, which contains location data in JSON-like format, @tbl-raw-coordinates displays how the data was like. A custom function was created to parse the coordinates, filter out invalid entries, and extract the longitude and latitude values. The extracted coordinates were combined into a new dataset shon in @tbl-clean-coordinates, with unique IDs to match the original entries. Any rows with missing coordinate data were removed, and the cleaned data was saved into a separate file for further analysis.

@@

Finally, the fourth cleaning process involved classifying cycling lanes based on the type of infrastructure. Data from the 'INFRA_HIGHORDER' and 'INFRA_LOWORDER' columns was cleaned, and missing values were removed. The lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, depending on their Bikeway Type data, detail of this classification can be found in @sec-app-comfort. The classification, along with year data for installation and upgrades, was saved for analysis on the types of cycling lanes and their comfort levels. @tbl-clean-comfort shows how it looks like.

@@

@fig-installed consists of two panels depicting the number of bikeways installed in Toronto by year, alongside a trend analysis of the installation pattern over time. @fig-installed-1 displays the raw count of bikeways installed per year from 2001 to 2024. A significant spike is observed in 2001, with over 400 bikeways installed in that year alone. This likely reflects a major investment in cycling infrastructure at the time. Following 2001, the number of installations drops sharply and stabilizes at much lower levels in subsequent years, with occasional small peaks around 2007, 2011, and 2021.

@@

@fig-upgraded illustrates the number of bikeways upgraded in Toronto from 2001 to 2023, with panel (a) showing the raw counts of upgrades and panel (b) visualizing the trend. @fig-upgraded-1 reveals that the most significant number of upgrades occurred in 2015, where nearly 100 bikeways were updated, marking a major peak. Outside of this spike, the number of annual upgrades remains relatively modest, typically fluctuating between 10 and 50 bikeways. Notable periods of increased upgrades are visible around 2012 and 2020, possibly reflecting concentrated efforts to modernize or expand existing cycling infrastructure in those years.

In @fig-upgraded-2, the overall trend is slightly upward, as indicated by the red trend line. Despite the notable peaks and valleys, the trend line suggests that, on average, the number of bikeway upgrades has increased over the years. The spike in 2015 significantly affects the trend, but even excluding that, the gradual rise suggests that more consistent efforts have been made to maintain and improve Toronto’s cycling network. This trend aligns with broader urban planning strategies focused on making cities more accessible and environmentally friendly by supporting active transportation. The dips in recent years, particularly after 2020, might be influenced by resource reallocation or external factors, such as the global pandemic, which could have delayed infrastructure projects.

@@

@fig-comfort-one-2, @fig-comfort-two-1, and @fig-comfort-two-2 offer a deeper look into the upgrade status for each comfort level. For Low Comfort lanes, only 22.26% have been upgraded, while 77.74% remain in their original state. Similarly, for Moderate Comfort bikeways, 23.38% have been upgraded, leaving 76.62% untouched. Interestingly, High Comfort lanes have the lowest proportion of upgrades, with only 18.76% improved, despite representing the most comfortable and potentially most desirable biking infrastructure. These trends highlight a potential gap in Toronto's focus on maintaining and upgrading its most comfortable and accessible cycling lanes, suggesting a need for more targeted investments in high-comfort bikeways to encourage broader use of cycling for transportation.

@@

@fig-map provides a visual representation of the bike lane infrastructure across Toronto. The lines on the map correspond to various cycling paths, including dedicated bike lanes, multi-use paths, and shared roadways, which span different areas of the city. The geographic coordinates, ranging from around 43.60°N to 43.85°N in latitude and -79.6°W to -79.3°W in longitude, encompass most of Toronto’s urban region. This map offers an important spatial perspective, illustrating the extensive network of cycling routes that connect different neighborhoods and areas within the city.

A key observation from the map is the dense concentration of bike lanes in the downtown core and along major roadways, indicating a well-established network in central Toronto. The downtown area’s grid-like structure is particularly evident, showcasing a high density of lanes that facilitate easy cycling access within the city's busiest areas. As one moves further away from the city center, the map shows fewer bike lanes, suggesting that suburban regions may have less developed cycling infrastructure. However, there are notable multi-use paths extending toward the outskirts, particularly in parks and recreational areas, providing important cycling connections for those commuting or using bikes for recreational purposes.

@@

The data reveals a phased development of Toronto's cycling infrastructure, with a notable increase in installations during the early 2000s, as shown in @fig-installed. This rise is likely due to the dataset commencing in 2001, thereby consolidating all installations prior to this year. Additionally, the onset of formal record-keeping and the significant number of installations during this period may reflect political momentum and public demand for environmentally sustainable transportation alternatives, influenced by growing concerns over climate change and urban congestion.

Ignoring the data from the first year, the annual number of new bikeway installations still showing a trend of slight decreasing. Several factors may account for this decline. In the early stages, the city's cycling plan likely targeted areas where infrastructure could be installed with relative ease, particularly in downtown Toronto, where road networks and urban density favored such investments. As the network extended into more complex regions, such as suburban areas with fewer existing roadways, the rate of installation slowed. Moreover, the physical limitations of space in a densely developed urban environment likely constrained further expansion, as new bikeways would require more complex and expensive measures, such as road reallocation or the creation of shared-use paths.

Despite this slowdown in new installations, @fig-upgraded demonstrates a substantial focus on upgrading existing infrastructure, particularly in recent years. This shift from new installations to upgrades may indicate a strategic decision by the city to enhance the quality and safety of the current cycling infrastructure. As the infrastructure ages and new safety standards are introduced, upgrading older lanes is essential to ensuring the network remains functional and safe for cyclists. The peak of upgrades in 2015, for instance, may reflect the city's response to increased cycling activity, advocacy group demands, or the adoption of updated cycling policies that prioritize safer designs, such as protected bike lanes.

The shift toward prioritizing upgrades over new installations also suggests that the initial expansion left certain areas with inadequate or unsafe infrastructure, particularly in the form of shared roadways or painted bike lanes offering minimal cyclist protection. As urban transportation planning evolves, the emphasis has shifted toward constructing more segregated and protected lanes, which are vital for encouraging cycling among less experienced riders, families, or individuals concerned about safety. Considering new plans by the goverment [@citeCyclingNetworkPlan], this ongoing focus on improving infrastructure—rather than simply expanding it—reflects a more advanced approach to active transportation, recognizing that safety and quality are equally important to increasing cycling participation.

In conclusion, the analysis highlights the progression of Toronto’s cycling infrastructure from an initial phase of rapid expansion to a more recent emphasis on upgrading and enhancing the safety and quality of the network. This evolution reflects the complexities of balancing growth with maintenance, the need to adapt to changing urban environments, and the influence of international best practices and local advocacy in shaping the future of cycling in Toronto.

@@

The classification of comfort levels in Toronto's cycling infrastructure offers significant insights into the city's capacity to ensure cyclist safety and encourage widespread usage. According to the Bikeway Comfort Level Distribution, @fig-comfort-one-1, a large portion of the network consists of lanes categorized as either High or Low Comfort, with Moderate Comfort lanes representing a smaller percentage. This distribution indicates that, although Toronto has expanded its cycling network, the infrastructure may still lack the protection required to encourage consistent cycling across a broader demographic. Ideally, Low Comfort lanes should form the smallest proportion, ensuring safer and more accessible routes for cyclists of all experience levels.

@@

In conclusion, while Toronto has made progress in expanding its cycling network, the high proportion of unupgraded lanes underscores the need for further improvements. The city should prioritize upgrading its Low and Moderate Comfort lanes while continuing to expand its High Comfort infrastructure. This shift will not only improve safety but also enhance equity in urban transportation, making cycling a more viable and appealing option for a diverse range of residents.

@@

The geospatial analysis of Toronto's cycling infrastructure, as shown in @fig-map, reveals both strengths and gaps in the city's bike network. One key observation is the dense concentration of bikeways in the downtown core. This concentration aligns with expectations, given that downtown Toronto is the city's most populated and congested area, with high demand for alternative transportation options. The extensive network of bikeways in this region reflects a targeted effort to make cycling a viable and accessible mode of transportation for residents, commuters, and visitors.

However, @fig-map also reveals a significant disparity in infrastructure coverage as one moves away from the city center toward suburban and peripheral areas. In these regions, cycling infrastructure is sparse, underscoring unequal access to safe and efficient bike routes. This limited infrastructure in suburban areas reduces the overall utility of the city's cycling network, as it disproportionately benefits those residing in the urban core. Suburban residents, often facing longer commutes and a greater dependence on automobiles, would benefit from an expanded bike lane network, providing a safer and more convenient alternative to driving.

Several factors contribute to this discrepancy in cycling infrastructure. Suburban areas typically have lower population densities and roadways designed primarily for car travel, which historically has deprioritized cycling in urban planning. Additionally, suburban cultures often rely heavily on automobiles, leading to lower public demand for bike lanes. As Toronto aims to promote sustainable transportation and reduce car dependency, prioritizing the development of safe cycling routes in suburban areas will be essential. Expanding infrastructure in these regions can significantly contribute to the city’s broader transportation and environmental goals.

@fig-map also emphasizes the importance of recreational areas, particularly those near the waterfront, in shaping the cycling network. These areas often feature multi-use trails with High Comfort lanes, offering cyclists scenic and safe routes away from motor traffic. These trails serve not only as commuting routes but also as recreational spaces, promoting exercise and relaxation.

Finally, @fig-map points to opportunities for future expansion. Areas with few or no cycling routes should be considered for infrastructure development, especially where there is sufficient road space or public demand for safer cycling options. As Toronto continues to grow, addressing these gaps in the cycling network will be critical to ensuring that all residents, regardless of location, have access to safe, comfortable, and convenient cycling infrastructure. Expanding the network equitably across the city will support the development of a more inclusive and sustainable transportation system.

In conclusion, while the spatial distribution of Toronto's cycling infrastructure demonstrates a well-connected downtown core, it also highlights the need for further investment in suburban regions. Expanding the cycling network in underserved areas will be crucial for promoting cycling as a practical and widely adopted mode of transportation throughout the city.

@@

One key limitation of this study is the reliance on the 'Cycling Network' dataset from OpenDataToronto, which, while comprehensive, might not capture informal or emerging cycling routes that are not officially recognized. Additionally, the dataset may not include newer infrastructure developments if updates were delayed in publication. Future research should integrate real-time data collection techniques or alternative datasets (e.g., crowdsourced information) to ensure that the analysis reflects the most current state of the network.

Another weakness lies in the classification of comfort levels, which, while useful, is a simplification. Comfort in cycling can be influenced by numerous factors beyond infrastructure type, such as traffic density, road conditions, and seasonal weather. A more nuanced classification that incorporates these additional factors could provide a more accurate assessment of how comfortable Toronto's cycling network truly is for different types of cyclists.

>>>>

---
title: "Geospatial and Temporal Analysis of Toronto's Cycling Infrastructure"
subtitle: "A Data-Driven Approach"
author: 
  - Ziheng Zhong
thanks: "Code and data are available at: https://github.com/iJustinn/Toronto_Cycling_Network"
date: today
date-format: long
abstract: "This study examines the evolution of Toronto's cycling infrastructure between 2001 and 2024, focusing on installation trends, upgrades, and comfort levels of bike lanes. The analysis reveals that while the cycling network has grown, many lanes remain low in comfort, with significant disparities in access between downtown and suburban areas. These findings highlight the need for continued investment in upgrading existing lanes and expanding the network to promote safer and more equitable cycling options across the city. Such improvements are essential for supporting sustainable transportation and reducing car dependency in Toronto."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# Package List
package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr", "here", "knitr", "kableExtra")

# Check and install missing packages
install_and_load <- function(package_list) {
  for (package in package_list) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

install_and_load(package_list)

@@

The development of urban cycling infrastructure is critical for promoting sustainable transportation, reducing greenhouse gas emissions, and improving public health. In rapidly growing cities like Toronto, the integration of cycling lanes into the urban transportation system provides an alternative to car dependency, addressing concerns related to traffic congestion and environmental sustainability. However, despite Toronto’s growing network of bike lanes, there remains significant variation in the quality and accessibility of these lanes, particularly between downtown and suburban areas. For most Toronto residents, they may see some bicycles riding on the road everyday, but there is a high probability that there is no clear understanding of Toronto's bike lanes construction. This paper is meant to fulfill this knowledge gap and give people an overall understanding on Toronto's cycling network development and why it is important.

This paper focuses on the evolution of Toronto’s cycling infrastructure from 2001 to 2024, providing a geospatial and temporal analysis of bikeway installations, upgrades, and comfort levels across the city. Using the publicly available 'Cycling Network' dataset [@citeCyclingNetwork] from Open Data Toronto [@citeODT], this study investigates the distribution of bike lanes, the different lanes' comfort levels with its classifications, and the trends in cycling infrastructure development over the past two decades. It highlights the differences in the network's coverage between the densely populated downtown core and the more sparsely served suburban regions.

A critical gap in the literature is the lack of detailed, data-driven studies assessing the quality of Toronto’s cycling infrastructure. While the city has made strides in expanding its network, many lanes remain of low comfort, which may discourage public choosing cycling as a daily mode of transportation. This paper aims to solve this problem by analyzing both the growth of the network and its current state in terms of usability and safety.

To address this, the study evaluates the installation and upgrade trends of bike lanes, classifies the comfort levels of these lanes, and identifies areas where further investment is necessary. The results indicate that while Toronto’s cycling network has expanded, many lanes offer minimal physical protection, with significant underinvestment in suburban areas. These findings underscore the need for continuous improvements to both the safety and accessibility of the network, which are essential for supporting Toronto’s broader goals of sustainable urban mobility.

The importance of this study lies in its potential to inform urban planners and policymakers about the current shortcomings of Toronto’s cycling infrastructure. By emphasizing the need for targeted investments, particularly in underserved regions, the study contributes to ongoing efforts to make cycling a more viable and equitable transportation option across the city.

The structure of this paper is organized as follows: @sec-data presents the data sources and methodologies, followed by an overview of the results in @sec-res. The findings are then discussed in @sec-dis, and additional detailed information is provided in the appendix at @sec-app.

@@

This project is structured based on Rohan's principles [@citeRohan]. Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages in R: `sf` [@citeSf], `readr` [@citeReadr], `ggplot2` [@citeGgplot2], `osmdata` [@citeOsmdata], `tidyverse` [@citeTidyverse], `jsonlite` [@citeJsonlite], `dplyr` [@citeDplyr], `here` [@citeHere], `knitr` [@citeKnitr], `kableExtra` [@citeKableExtra].

## Source
The dataset analyzed in this study, the Cycling Network dataset [@citeCyclingNetwork], was retrieved from Open Data Toronto [@citeODT]. This dataset is publicly available and is continually updated to reflect Toronto’s ongoing efforts to develop and maintain its cycling infrastructure. It contains geospatial and temporal information about various cycling-related features, including the types of bikeways, installation years, etc. Such a detailed history of Toronto's cycling network development shows a good fit for this study.

While the Cycling Network dataset was the primary source of data, other relevant datasets were explored, such as the Road Network File [@citeStatCan] from Statistics Canada, which provides detailed geospatial data on all roadways in Toronto. Another alternative was the Major City-Wide Cycling Routes dataset [@citeCyclingRoutes], also available on Open Data Toronto [@citeODT]. However, this dataset focuses specifically on the large-scale cycling routes connecting different parts of the city, emphasizing connectivity at a macro level. While these two dataset are useful for analyzing high-level transportation patterns, it lacks the detailed data such as comfort levels, installation dates, and upgrade history, that are crucial for a deep analysis of cycling infrastructure. Since this paper focuses on both the growth and quality of cycling infrastructure, the Cycling Network dataset, with its more comprehensive coverage of individual bikeway segments, was a better fit for the goals of this study.

@@

The 'Cycling Network' dataset from Open Data Toronto tracks various aspects of Toronto's cycling infrastructure, such as installation dates, upgrades, street names, and the type of the lane. Each entry represents a segment of the cycling network, such as the construction or upgrading of cycling paths, are documented. For example, when a cycling path is installed or upgraded, the responsible authorities collect information such as the installation year and any year of subsequent upgrades. This data is then inputed into structured entries in the dataset.

To go from a physical event, such as installing a new lane, to a dataset entry, detailed records are maintained by the city's transportation department. These records include precise latitude and longitude of the lanes, allowing each cycling path to be mapped precisely in relation to other city infrastructure. Thus, the phenomena of constructing or updating a bike lane becomes an entry with attributes such as the street name, road class, and installation history. This allows researchers to analyze trends in cycling infrastructure and its expansion over time.

@@

The various variables mentioned above serves as key variables that were used in analysis for this paper:

@@

#### Comfort Level

Each bikeway is assigned a comfort level—high, moderate, or low—based according to the recorded Bikeway Type. These classification are self-defined, reling on factors such as physical separation from traffic and lane width of each Bikeway Type, detail can be found in @sec-app-comfort. This variable is central to understanding the usability of the infrastructure.

@@

@@

In the first data cleaning process, the focus was on extracting and cleaning geographical coordinates from the 'geometry' column, which all data are in JSON-like format, @tbl-raw-coordinates displays how the raw data was like. A function was created to parse the coordinates, filter out invalid entries, and extract the longitude and latitude values. The extracted coordinates were combined into a new dataset shown in @tbl-clean-coordinates, with unique IDs to match the original cycling lanes they from. Any rows with missing coordinate data were removed, and the cleaned data was saved into a separate file for further analysis.

@@

Finally, the fourth cleaning process involved classifying cycling lanes based on the type of infrastructure. Data from the 'INFRA_HIGHORDER' and 'INFRA_LOWORDER' columns was cleaned, and missing values were removed. By using the Bikeway Type data, the lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, detail of this classification can be found in @sec-app-comfort. The classification, along with year data for installation and upgrades, was saved for analysis on the types of cycling lanes and their comfort levels. @tbl-clean-comfort shows how it looks like.

@@

@fig-installed consists of two panels depicting the number of bikeways installed in Toronto by year, alongside a trend analysis of the installation pattern over time. @fig-installed-1 displays the raw count of bikeways installed per year from 2001 to 2024. A significant spike is observed in 2001, with over 400 bikeways installed in that year alone. This likely reflects the starting time of this dataset, which all lanes built before 2001 was recorded as 2001. Following 2001, the number of installations drops sharply and stabilizes at much lower levels in subsequent years, with occasional small peaks around 2007, 2011, and 2021.

@@

@fig-upgraded illustrates the number of bikeways upgraded in Toronto from 2001 to 2024, with panel (a) showing the raw counts of upgrades and panel (b) visualizing the trend. @fig-upgraded-1 reveals that the most significant number of upgrades occurred in 2015, where nearly 100 bikeways were updated, marking a major peak. Outside of this spike, the number of annual upgrades remains relatively modest, typically fluctuating between 10 and 50 bikeways each year. Notable periods of increased upgrades are visible around 2012 and 2020, possibly reflecting concentrated efforts to modernize or expand existing cycling infrastructure in those years.

In @fig-upgraded-2, the overall trend is slightly upward, as indicated by the red trend line. Despite the notable peaks and drops, the trend line suggests that, on average, the number of bikeway upgrades has increased over the years. The spike in 2015 significantly affects the trend, but even excluding that, the gradual rise suggests that more consistent efforts have been made to maintain and improve Toronto’s cycling network. This trend aligns with broader urban planning strategies focused on making cities more accessible and environmentally friendly by supporting active transportation. The dips in recent years, particularly after 2020, might be influenced by resource reallocation or external factors, such as the global pandemic, which could have delayed some plans.

@@

@fig-comfort-one-2, @fig-comfort-two-1, and @fig-comfort-two-2 offer a deeper look into the upgrade status for each comfort level. For Low Comfort lanes, only 22.26% have been upgraded, while 77.74% remain in their original state. Similarly, for Moderate Comfort bikeways, 23.38% have been upgraded, leaving 76.62% untouched. Interestingly, High Comfort lanes have the lowest proportion of upgrades, with only 18.76% improved. This result shows not much cycling lanes achieved high comfort level by upgrading, most of them are initially designed and built to be high comfort lanes. These trends highlight a potential gap in Toronto's focus on maintaining and upgrading its most comfortable and accessible cycling lanes, suggesting a need for more targeted investments in high-comfort bikeways to encourage cycling in the city.

@@

@fig-map provides a visual representation of the bike lane infrastructure across Toronto. The lines on the map correspond to various cycling paths, which span different areas of the city. The geographic coordinates, ranging from around 43.60°N to 43.85°N in latitude and -79.6°W to -79.3°W in longitude, encompass most of Toronto’s urban region. This map offers an important spatial perspective, illustrating the extensive network of cycling routes that connect different neighborhoods and areas within the city.

A key observation from the map is the dense concentration of bike lanes in the downtown areas and along major roadways, indicating a well-established network in central Toronto. The downtown area’s grid-like structure is particularly evident, showcasing a high density of lanes that facilitate easy cycling access within the city's busiest areas. As one moves further away from the city center, the map shows fewer bike lanes, suggesting that suburban regions may have less developed cycling infrastructure.

@@

The data reveals a phased development of Toronto's cycling infrastructure, with a notable increase in installations during the early 2000s, as shown in @fig-installed. As mentioned before, this rise is likely due to the dataset starting in 2001, thereby recorded as 2001 for all installations prior to this year. However, if there was a large numebr of lanes installed in 2001, then it may reflect political momentum and public demand for environmentally sustainable transportation alternatives, influenced by growing concerns over climate change and urban congestion.

Ignoring the data from the first year, the annual number of new bikeway installations will still show a trend of slight decreasing. Several factors may account for this decline. In the early stages, the city's cycling plan likely targeted areas where infrastructure could be installed with relative low difficulties, particularly in downtown Toronto, where road networks and urban density favored such investments. As the network extended into more complex regions, such as suburban areas with fewer existing roadways, the rate of installation slowed. Moreover, the physical limitations of space in a densely developed urban environment likely constrained further expansion, as new bikeways would require more complex and expensive measures, such as road reallocation or the creation of shared-use paths.

Despite this slowdown in new installations, @fig-upgraded demonstrates a substantial focus on upgrading existing infrastructure, particularly in recent years. This shift from new installations to upgrades may indicate a strategic decision by the city to enhance the quality and safety of the current cycling infrastructure. As the infrastructure ages and new safety standards are introduced, upgrading older lanes is essential to ensuring the network remains functional and safe for cyclists. Also, some study have shown the possibilities of improving traffic issues by upgrading the existing cycling lanes [@citeCyclingImprove], this study also mentioned increasing the accessibility can highly impact people's will to choose cycling to commute.

The shift toward prioritizing upgrades over new installations also suggests that the initial expansion left certain areas with inadequate or unsafe infrastructure, particularly in the form of shared roadways or painted bike lanes offering minimal cyclist protection. As urban transportation planning evolves, the emphasis has shifted toward constructing more segregated and protected lanes, which are vital for encouraging cycling among less experienced riders or individuals concerned about safety. Considering new plans by the goverment [@citeCyclingNetworkPlan], this ongoing focus on improving infrastructure rather than simply expanding it, reflects a more advanced approach to active transportation, recognizing that safety and quality are equally important to increasing cycling participation.

@@

@@

Moreover, a study have shown the importance of buffer zones in cycling lanes [@citeLaneBuffer], it not only provide higher safety level, but also give people higer confidence to use these lanes when there are vehicles on the road. According to the Bikeway Comfort Level Distribution, @fig-comfort-one-1, a large portion of the network consists of lanes categorized as either High or Low Comfort, with Moderate Comfort lanes representing the samllest percentage. This distribution indicates that, although Toronto has expanded its cycling network, the infrastructure may still lack the high quality cycling lanes to encourage consistent cycling across mroe people. Ideally, Low Comfort lanes should form the smallest proportion, ensuring safer and more accessible routes for cyclists of all experience levels. In such cases, combining what we saw in the buffer zone study, there will be more usage of cycling lanes, thus help Toronto become a healthier and more environmental friendly city.

@@

The geospatial analysis of Toronto's cycling infrastructure, as shown in @fig-map, reveals both strengths and gaps in the city's bike network. One key observation is the dense concentration of bikeways in the downtown Toronto. This concentration aligns with expectations, given that downtown areas are the city's most populated and crowded area, with high demand for alternative transportation options. The extensive network of bikeways in this region reflects a targeted effort to make cycling a viable and accessible mode of transportation for residents and visitors.

However, @fig-map also reveals a significant disparity in infrastructure coverage as one moves away from the city center toward suburban areas. In these regions, cycling infrastructure is sparse, underscoring unequal access to safe and efficient bike routes. This limited infrastructure in suburban areas reduces the overall utility of the city's cycling network, as it cannot connect to all areas that you may want to go. Several factors contribute to this discrepancy in cycling infrastructure. Suburban areas typically have lower population densities and roadways designed primarily for car travel, which historically has lowered the priority of cycling infrastructure. However, expand cycling lanes to these areas would still benefit people since it provids an alternative choice other than driving. This can also help those who haven't get a driver's license to have more flexible ways of transportation. As Toronto aims to promote sustainable transportation and reduce car dependency, prioritizing the development of safe cycling routes in suburban areas will be essential. Expanding infrastructure in these regions can significantly contribute to the city’s broader transportation and environmental goals [@citeEnvGoal].

@fig-map also emphasizes the importance of recreational areas, particularly those near the waterfront, in shaping the cycling network. These areas often feature multi use trails with High Comfort lanes, offering cyclists scenic and safe routes away from motor traffic. These trails serve not only as commuting routes but also as recreational spaces, promoting exercise and relaxation.

Finally, @fig-map points to opportunities for future expansion. Areas with few or no cycling routes should be considered for infrastructure development, especially where there is sufficient road space or public demand for safer cycling options. As Toronto continues to grow, addressing these gaps in the cycling network will be critical to ensuring that all residents, regardless of location, have access to safe, comfortable, and convenient cycling infrastructure. Lastly, it was mentioned as the first choice in the 'Low Carbon Transportation' article [@citeLowCarb] by City of Toronto, it further emphasized the importance of axpanding the network across the city to support the development of a more inclusive and sustainable transportation system.

@@

One key limitation of this study is the reliance on the 'Cycling Network' dataset from Open Data Toronto, which, while comprehensive, might not capture informal or emerging cycling routes that are not officially recognized. Additionally, the dataset may not include newer infrastructure developments if updates were delayed in publication. Future research should integrate real-time data collection techniques or alternative datasets such as crowdsourced information to ensure that the analysis reflects the most current state of the network.

Another weakness lies in the classification of comfort levels, which is  useful but is an over simplification of this problem. Comfort in cycling can be influenced by numerous factors beyond infrastructure type, such as traffic density, road conditions, etc. A more comprehensive classification that incorporates these additional factors could provide a more accurate assessment of how good Toronto's cycling network truly is.
